Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-09-07 paper.Rnw
small updates to improve pcp

# Extract the features and order them based on feature importance
bullet_features <- rownames(bulletxtrctr::rtrees$importance)
bullet_features_ordered <-
  data.frame(
    feature = rownames(bulletxtrctr::rtrees$importance),
    MeanDecreaseGini = bulletxtrctr::rtrees$importance
  ) %>%
  arrange(desc(MeanDecreaseGini)) %>%
  mutate(
    feature = fct_recode(
      feature,
      "CCF" = "ccf",
      "CMS" = "cms",
      "Matches" = "matches",
      "Mismatches" = "mismatches",
      "Non-CMS" = "non_cms",
      "Rough Correlation" = "rough_cor",
      "Distance Std Dev" = "sd_D",
      "Distance" = "D",
      "Sum Peaks" = "sum_peaks"
    )
  )

@@

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes \Sexpr{dim(bullet_train)[1]} land-to-land comparisons.

@@

# Create plot of the signatures
signatures %>%
  mutate(land = c("Signature 1", "Signature 2")[as.factor(source)]) %>%
  ggplot(aes(x = x/1000)) + 
  geom_line(aes(y = sig), colour = "grey30") +
  facet_grid(land~.) +
  ylim(c(-4,6)) +
  theme_bw(base_family = ff, base_size = f9_fs) +
  theme(strip.background = element_blank()) +
  ylab("Surface measurement (in microns)") +
  xlab("Relative Location (in millimeters)")

@@

# Create (or load) the parallel coordinate plot
if (!file.exists("./figure-static/figure-10-1.png")) {

@@

  # Create the plot
  bullet_pcp <- bullet_train %>%
    select(all_of(bullet_features), samesource, rfscore) %>%
    bind_rows(bullet_test %>%
                select(all_of(bullet_features), samesource, rfscore),
              .id = "set") %>%
    mutate(
      set = fct_recode(set,
                       "Training Set" = "1",
                       "Testing Set" = "2"),
      samesource = fct_recode(
        factor(samesource),
        "Match" = "TRUE",
        "Non-Match" = "FALSE"
      )
    ) %>%
    rename(
      "CCF" = "ccf",
      "CMS" = "cms",
      "Matches" = "matches",
      "Mismatches" = "mismatches",
      "Non-CMS" = "non_cms",
      "Rough Correlation" = "rough_cor",
      "Distance Std Dev" = "sd_D",
      "Distance" = "D",
      "Sum Peaks" = "sum_peaks"
    ) %>%
    na.omit() %>%
    ggplot(aes(color = rfscore)) +
    geom_pcp(
      mapping = aes(vars = vars(bullet_features_ordered$feature)),
      alpha = 0.2,
      size = f10_ls
    ) +
    facet_grid(set ~ samesource) +
    scale_color_gradient2(low = "grey50",
                          high = "darkorange",
                          midpoint = 0.5) +
    theme_bw(base_family = ff, base_size = f10_fs) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.placement = "outside",
      strip.background = element_rect(color = "white",
                                      fill = "white")
    ) +
    labs(x = "Features Ordered by Random Forest Variable Importance \n(highest to lowest from left to right)",
         y = "Scaled Feature Values",
         color = "Random \nForest \nProbability")

@@

# Create (or load) the parallel coordinate plot
if (file.exists("./figure-static/figure-10-1.eps")) {

@@

# Adjust names of the cases
bullet_explain_perms <- bullet_explain_perms %>%
  mutate(case = ifelse(case == bullet_poi_nonmatch, "NM", "M"))

# Create the explanation scatterplot: case M, 3 quantile bins
eoiM_3qb <-
  plot_explain_scatter(
    bullet_explain_perms[1:3,],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = ff, base_size = f13_fs) +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(size = f13_fs),
    strip.text.x = element_text(size = f13_fs),
    strip.text.y = element_text(size = f13_fs),
    strip.placement = "outside",
    strip.background = element_rect(color = "white", fill = "white")
  ) +
  guides(linetype = guide_legend(override.aes = list(color = c("darkorange"))),
         size = guide_legend(nrow = 2, byrow = T))

# Create the explanation scatterplot: case MM, 3 quantile bins
eoiNM_3qb <-
  plot_explain_scatter(
    bullet_explain_perms[4:6, ],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = ff, base_size = f13_fs) +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(size = f13_fs),
    strip.placement = "outside",
    strip.background = element_rect(color = "white", fill = "white"),
    strip.text.x = element_text(size = f13_fs),
    strip.text.y = element_text(size = f13_fs)
  ) +
  guides(linetype = guide_legend(override.aes = list(color = c(
    "grey50", "darkorange"
  ))),
  size = guide_legend(nrow = 2, byrow = T))

# Create the explanation scatterplot: case M, 3 equal bins
eoiM_3eb <-
  plot_explain_scatter(
    bullet_explain_perms[7:9,],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = ff, base_size = f13_fs) +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(size = f13_fs),
    strip.placement = "outside",
    strip.background = element_rect(color = "white", fill = "white"),
    strip.text.x = element_text(size = f13_fs),
    strip.text.y = element_text(size = f13_fs)
  ) +
  guides(linetype = guide_legend(override.aes = list(color = c(
    "grey50", "darkorange"
  ))),
  size = guide_legend(nrow = 2, byrow = T))

# Create the explanation scatterplot: case NM, 3 equal bins
eoiNM_3eb <-
  plot_explain_scatter(
    bullet_explain_perms[10:12,],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  scale_color_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  scale_fill_gradient2(
    low = "grey50",
    high = "darkorange",
    midpoint = 0.5,
    limits = c(0, 1)
  ) +
  theme_bw(base_family = ff, base_size = f13_fs) +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(size = f13_fs),
    strip.placement = "outside",
    strip.background = element_rect(color = "white", fill = "white"),
    strip.text.x = element_text(size = f13_fs),
    strip.text.y = element_text(size = f13_fs)
  ) +
  guides(linetype = guide_legend(override.aes = list(color = c(
    "grey50", "darkorange"
  ))),
  size = guide_legend(nrow = 2, byrow = T))

# Create the lime explanation visualization: case M, 3 quantile bins
pfM_3qb <- plot_features(bullet_explain_perms[1:3,]) +
  scale_fill_manual(values = c("darkorange", "grey50")) +
  theme(text = element_text(family = ff, size = f13_fs),
        strip.text.x = element_text(face = "plain", size = f13_fs))

# Create the lime explanation visualization: case NM, 3 quantile bins
pfNM_3qb <- plot_features(bullet_explain_perms[4:6,]) +
  scale_fill_manual(values = c("darkorange", "grey50")) + 
  theme(text = element_text(family = ff, size = f13_fs),
        strip.text.x = element_text(face = "plain", size = f13_fs))

# Create the lime explanation visualization: case M, 3 equal bins
pfM_3eb <- plot_features(bullet_explain_perms[7:9,]) +
  scale_fill_manual(values = c("darkorange", "grey50")) +
  theme(text = element_text(family = ff, size = f13_fs),
        strip.text.x = element_text(face = "plain", size = f13_fs))

# Create the lime explanation visualization: case NM, 3 equal bins
pfNM_3eb <- plot_features(bullet_explain_perms[10:12,]) +
  scale_fill_manual(values = c("darkorange", "grey50")) +
  theme(text = element_text(family = ff, size = f13_fs),
        strip.text.x = element_text(face = "plain", size = f13_fs))

@@

The following supporting information is available:

\kge{\begin{itemize}
\item \textbf{Paper GitHub Repository}: The R markdown document and code used to produce the paper are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper-short-sweet}{https://github.com/goodekat/LIME-diagnostics-paper-short-sweet}. \kgc{Need to update this link after I change the name of the repository.}
\item \textbf{Data}: The data used for the examples in Section \ref{application} are available at \kgc{need to fill in once Heike has the data published.}
\end{itemize}}

>>>>

# Extract the features and order them based on feature importance
bullet_features <- rownames(bulletxtrctr::rtrees$importance)
bullet_features_ordered <-
  data.frame(
    feature = rownames(bulletxtrctr::rtrees$importance),
    MeanDecreaseGini = bulletxtrctr::rtrees$importance
  ) %>%
  arrange(desc(MeanDecreaseGini)) %>%
  mutate(
    feature = fct_recode(feature,
      "Cross Correlation Function" = "ccf",
      "Consecutively Matching Striae" = "cms",
      "Matches" = "matches",
      "Mismatches" = "mismatches",
      "Non-Consecutively Matching Striae" = "non_cms",
      "Rough Correlation" = "rough_cor",
      "Distance Standard Deviation" = "sd_D",
      "Distance" = "D",
      "Sum of Peaks" = "sum_peaks"
    )
  )

@@

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). \kge{See the supporting information for additional information on the data.} The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes \Sexpr{dim(bullet_train)[1]} land-to-land comparisons.

@@

# Create plot of the signatures
signatures %>%
  mutate(land = c("Signature 1", "Signature 2")[as.factor(source)]) %>%
  ggplot(aes(x = x/1000)) + 
  geom_line(aes(y = sig), colour = "grey30") +
  facet_grid(land~.) +
  ylim(c(-4,6)) +
  theme_bw(base_family = ff, base_size = f9_fs) +
  theme(strip.background = element_blank()) +
  ylab("Surface Measurement (in microns)") +
  xlab("Relative Location (in millimeters)")

@@

# Create or load the parallel coordinate plot
if (!file.exists("./figure-static/figure-10-1.png")) {

@@

  # Create the plot
  bullet_pcp <- bullet_train %>%
    select(all_of(bullet_features), samesource, rfscore) %>%
    bind_rows(bullet_test %>%
                select(all_of(bullet_features), samesource, rfscore),
              .id = "set") %>%
    mutate(
      set = fct_recode(set,
                       "Training Set" = "1",
                       "Testing Set" = "2"),
      samesource = fct_recode(
        factor(samesource),
        "Match" = "TRUE",
        "Non-Match" = "FALSE"
      )
    ) %>%
    rename(
      "Cross Correlation Function" = "ccf",
      "Consecutively Matching Striae" = "cms",
      "Matches" = "matches",
      "Mismatches" = "mismatches",
      "Non-Consecutively Matching Striae" = "non_cms",
      "Rough Correlation" = "rough_cor",
      "Distance Standard Deviation" = "sd_D",
      "Distance" = "D",
      "Sum of Peaks" = "sum_peaks"
    ) %>%
    na.omit() %>%
    ggplot(aes(color = rfscore)) +
    geom_pcp(
      mapping = aes(vars = vars(bullet_features_ordered$feature)),
      alpha = 0.2,
      size = f10_ls
    ) +
    facet_grid(set ~ samesource) +
    scale_color_gradient2(low = "grey50",
                          high = "darkorange",
                          midpoint = 0.5) +
    theme_bw(base_family = ff, base_size = f10_fs) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.placement = "outside",
      strip.background = element_rect(color = "white",
                                      fill = "white")
    ) +
    labs(x = "Features Ordered by Random Forest Variable Importance \n(highest to lowest from left to right)",
         y = "Scaled Feature Values",
         color = "Random \nForest \nProbability")

@@

@@

# Adjust names of the cases
bullet_explain_perms_clean <- bullet_explain_perms %>%
  mutate(case = ifelse(case == bullet_poi_nonmatch, "NM", "M"))

# Create a theme for the lime explanation visualizations
bullet_explain_plot_theme <-
  list(
    scale_fill_manual(values = c("darkorange", "grey50")),
    theme(
      text = element_text(family = ff, size = f13_fs),
      strip.text.x = element_text(face = "plain", size = f13_fs)
    )
  )

# Create the lime explanation visualizations
pfM_3qb <- plot_features(bullet_explain_perms_clean[1:3,]) + bullet_explain_plot_theme
pfNM_3qb <- plot_features(bullet_explain_perms_clean[4:6,]) + bullet_explain_plot_theme
pfM_3eb <- plot_features(bullet_explain_perms_clean[7:9,]) + bullet_explain_plot_theme
pfNM_3eb <- plot_features(bullet_explain_perms_clean[10:12,]) + bullet_explain_plot_theme

# Create theme for explanation scatterplots
bullet_eoi_plot_theme <-
  list(
    scale_color_gradient2(
      low = "grey50",
      high = "darkorange",
      midpoint = 0.5,
      limits = c(0, 1)
    ),
    scale_fill_gradient2(
      low = "grey50",
      high = "darkorange",
      midpoint = 0.5,
      limits = c(0, 1)
    ),
    theme_bw(base_family = ff, base_size = f13_fs),
    theme(
      aspect.ratio = 1,
      plot.title = element_text(size = f13_fs),
      strip.text.x = element_text(size = f13_fs),
      strip.text.y = element_text(size = f13_fs),
      strip.placement = "outside",
      strip.background = element_rect(color = "white", fill = "white")
    ),
    guides(size = guide_legend(nrow = 2, byrow = T), 
           linetype = guide_legend(override.aes = list(color = c("grey50", "darkorange"))))
  )

# Create the explanation scatterplots
eoiM_3qb <-
  plot_explain_scatter(
    bullet_explain_perms_clean[1:3, ],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  bullet_eoi_plot_theme +
  guides(linetype = guide_legend(override.aes = list(color = c("darkorange"))))
eoiNM_3qb <-
  plot_explain_scatter(
    bullet_explain_perms_clean[4:6, ],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  bullet_eoi_plot_theme
eoiM_3eb <-
  plot_explain_scatter(
    bullet_explain_perms_clean[7:9, ],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  bullet_eoi_plot_theme
eoiNM_3eb <-
  plot_explain_scatter(
    bullet_explain_perms_clean[10:12,],
    alpha = 0.9,
    weights = TRUE,
    line_size = f13_ls
  ) +
  bullet_eoi_plot_theme

@@

\kge{The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}. Additional information about the bullet matching data are available at \kgc{XXX}.}

### 2020-09-07 paper.tex
small updates to improve pcp

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes 83028 land-to-land comparisons.

@@

%The following supporting information is available as part of the online article:

>>>>

In response, \citet{hare:2016} proposed an automated machine learning method for bullet matching to complement a visual inspection by firearm examiners. Based on high-resolution topological scans of land engraved areas, \citet{hare:2016} obtain signatures of striations from two bullet lands (\autoref{fig:figure-09}). Nine features quantifying the similarity of signatures, such as the cross-correlation function, the distance between signatures, and the number of matching striae, are extracted and used to train a random forest (available in the \emph{bulletxtrctr} R package) to determine the probability of a comparison resulting from the same source (matching signatures) or from different sources (non-matching signatures). \kge{See the supporting information for additional information on the data.} The model was trained on a set of scans of bullets from the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}, which includes 83028 land-to-land comparisons.

@@

\kge{The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}. Additional information about the bullet matching data are available at \kgc{XXX}.}
